---
title: "Springboard Capstone Project - CEI"
author: "James Hamilton"
date: "March 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(lubridate)
```

# The Question

What useful insights can be drawn from the clients' customer and invoice data?
    - Are some clients "over purchasing" because they don't understand the licensing scheme?
    - What market segments are the largest / smallest and which market segments are growing / shrinking?
    
# The Data
```{r load the data, echo=FALSE}
accounts = read.csv("SB-CEI-Capstone-Account.csv", header = TRUE)
orders = read.csv("SB-CEI-Capstone-Order.csv", header = TRUE)
orderlines = read.csv("SB-CEI-Capstone-Order-Product-Line.csv", header=TRUE)
```

The data was provided by the client in the form of a text CSV export from their Salesforce database.

I focused primarily on the Account, Order, and Order Line tables. Some cleanup of the data was needed - a detailing of that activity is included at the end of this report.

### Account

data included the customer name and demographic data: location, employee counts, annual revenue and various industry classifiers.

A correlation test indicates that the number of employees and Annual revenue are highly correlated (.93 at 99% confidence) across all customers indicating that using employee count alone as an indicator of the "size" of the customer would be accurate. Therefore, the following visuals are based on that conclusion and use the employee count as the primary indicator of client size.
```{r Accounts, echo=FALSE}
accounts.w <- accounts %>% filter(!is.na(accounts$Employees) & 
    accounts$Employees > 0 & 
    !is.na(accounts$Annual.Revenue) & 
    accounts$Annual.Revenue > 0)

p1 <- ggplot(data = accounts.w, aes(x = Industry)) + 
  geom_bar() +
  theme(axis.text.x = element_text(size=10, angle=35, hjust=1)) +
  ylab("Count of customers")

p2 <- ggplot(data = accounts.w, aes(Industry, Employees)) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,1000)) +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ylab("Employees")

grid.arrange(p2, p1)
```

Primary conclusions here are that the Machinery, Manufacturing and Construction industries represent the majority of the client base and those companies are predominately < 250 employees in size. "Other", "Retail", "Chemicals", "Engineering", and "Consulting" are also fairly large and will be studied as well.

```{r echo=FALSE}
Industries <- c("Chemicals","Construction","Consulting","Engineering","Machinery","Manufacturing","Other","Retail")

filter(accounts.w, Industry %in% Industries) %>%
  group_by(Industry) %>%
  select(Employees) %>%
  summary(.)
```

Here we see that these industries are represented by companies that are generally in the range of 0 - 140 employees. Specifically, across all industries about 75% of all clients have <= 140 employees. This does not however, describe which industries or companies are the "most valuable" in terms of dollars spent.


### Orders

data spanned two tables and included primarily dollar amounts, dates, quantities, payment methods and keys into other tables including the invoices and accounts tables. We are looking for two key pieces of information from this data:
    - The ratio of customer size to the amount of purchase made by industry. (in dollars and quantity of software licenses or seats purchased) This will be used to identify outliers.
    - The rate of change in amount of purchase made per industry
```{r Orders, echo=FALSE}
# Isolating columns needed from the three tables
orders$date.of.order <- mdy(orders$Order.Date)
working.ol <- orderlines[ , c("SalesOrder","Product","ProductFamily","ProductAmount","Quantity") ]
working.orders <- orders[ , c("Record.ID","date.of.order","Order.Type","Account.Number") ]
working.orders <- rename(working.orders, SalesOrder = Record.ID)
working.acct <- accounts[ , c("Account.Number",         "Billing.Country",
                              "Billing.State.Province", "Employees",
                              "Industry",               "Ownership")]
work.o <- left_join(working.ol, working.orders, by = "SalesOrder")
work <- left_join(work.o, working.acct, by = "Account.Number")
work <- filter(work, Industry %in% Industries)

byindustry <- group_by(work, Industry) %>%
  summarise(median_e_count = median(Employees),
    tot_e = sum(Employees),
    median_quantity = median(Quantity),
    mean_quantity = mean(Quantity),
    max_quantity = max(Quantity),
    tot_quantity = sum(Quantity),
    median_dollars = median(ProductAmount),
    mean_dollars = mean(ProductAmount),
    max_dollars = max(ProductAmount),
    tot_dollars = sum(ProductAmount)
    )

# I chose the mean for the quantity b/c the median for all industries is 1.
byindustry$qtyratio <- byindustry$mean_quantity / byindustry$median_e_count
byindustry$dollaratio <- byindustry$median_dollars / byindustry$median_e_count

ggplot(data = byindustry, aes(x = Industry)) +
  geom_point(aes(y = dollaratio, size = qtyratio, color = tot_dollars)) +
  coord_flip()

byaccount <- filter(work, !is.na(Account.Number) & !is.na(Employees) & Employees > 0) %>%
  group_by(., Industry, Account.Number, SalesOrder, date.of.order, Employees) %>%
  summarise(., 
    total_order_quantity = sum(Quantity),
    total_order_dollars = sum(ProductAmount)
  ) %>%
  arrange(., date.of.order)
byaccount$qty_per_emp <- byaccount$total_order_quantity / byaccount$Employees
byaccount$dol_per_emp <- byaccount$total_order_dollars / byaccount$Employees
byaccount$year.of.order <- year(byaccount$date.of.order)

ggplot(byaccount, aes(x = date.of.order)) +
  geom_point(aes(y = qty_per_emp, color = dol_per_emp)) +
  facet_wrap(~ Industry) + 
  xlab("Date")
```

It is important to note that about 1300 orders were removed from this analysis due to missing important column info - primarily Employee count.

Looking at these plots it appears that none of the Industries represented have a large group of purchases that are above the overall median of 1. This should be good news because that would suggest that the majority of clients have not overbought.

However, the display is fairly crowded and doesn't include the baseline value per industry so I'll focus on those industry groupings that appear to have the largest number of clients / orders and show a plot for each. Specifically: Chemicals, Construction, Consulting, Engineering, Machinery, Manufacturing, Other, and Retail.

These plots show several interesting pieces of data. First, each point is an individual order and is plotted by order date and the quantity of the order normalized by employee count. The total size of the order governs the color of the point and the dollars spent per employee dictates the size of each point. Larger and "cooler" dots should be cause for concern if they fall above the overall median.

The dashed horizontal line represents the ratio of the mean order quantity to the median employee count. The solid horizontal line represents the overall median order quantity across all clients.

The expectation is that half of the points would fall below the dashed line and that most of the points would fall below the solid line - indicating that most purchases are within expectations if clients are purchasing only what they need.

The data points above the line in all the critical industries are relatively small in number indicating a relatively low risk for those clients about which we have demographic data.

It looks as though for most industries the majority of purchases would be characterized as "correct."

```{r Industry Specific, echo = FALSE, message=FALSE}
plot.Industry <- function(i) {
  ratio <- filter(byindustry, Industry == i)$qtyratio
  
  ggplot(data = subset(byaccount, Industry == i),
    aes(x = date.of.order)) +
    geom_point(aes(y = qty_per_emp, size = dol_per_emp, color = total_order_dollars)) +
    scale_color_gradientn(colors=rainbow(7)) +
    geom_line(aes(y = ratio, group=1), linetype="dashed", color="black") +
    geom_line(aes(y = 1, group=1), linetype="solid", color="black") +
    theme(axis.text.x = element_text(size=10, angle=35, hjust=1)) +
    xlab("Order Date") +
    ylab("Order Qty per Employee") +
    ggtitle(i)
}

lapply(Industries, plot.Industry)
```

### How can we quantify the risk?

Specifically, how much of a problem could there be if the accounts that may be over purchased were to normalize their purchasing? How much of a risk is represented by the accounts that were excluded from the plots due to missing demographics?

Calculating the sum of the orders that lie above the overall median Order quantity per employee for the most recent full year should give a decent idea of the overall risk.
```{r risk}

```


```{r 2015 Total, echo=FALSE}
orders.2015 <- filter(byaccount, year.of.order == 2015)
q3.2015 <- quantile(byaccount$total_order_dollars, probs = 0.75)

dollars.2015 <- orders.2015 %>% ungroup() %>% select(total_order_dollars) %>% sum(.)
q4.dollars.2015 <- filter(orders.2015, total_order_dollars > q3.2015) %>%
  ungroup() %>%
  select(total_order_dollars) %>%
  sum(.)

q4.projected.2015 <- filter(orders.2015, total_order_dollars > q3.2015) %>% count(.)
q4.projected.2015 <- q4.projected.2015 * median(byaccount$total_order_dollars)

industry.totals.2015 <- filter(byaccount, year.of.order == 2015) %>%
  group_by(., Industry) %>%
  summarise(., 
    median.order = median(total_order_dollars),
    q4 = quantile(total_order_dollars, probs = 0.75),
    q4.count = sum(total_order_dollars > q4),
    allorders = sum(total_order_dollars),
    q4.total = sum(ifelse(total_order_dollars > q4, total_order_dollars, 0)),
    q4.projected = sum(ifelse(total_order_dollars > q4, median.order, 0))
  )

industry.totals.2015$delta <- with(industry.totals.2015, q4.total - q4.projected)

```

The 4th Quartile for orders in 2015 starts at $
`r q3.2015`   
Total dollar value of orders in 2015: $`r as.character(dollars.2015)`   
Total dollar value of orders in the 4th quartile: $`r as.character(q4.dollars.2015)`   

Projected revenue decrease if orders in 4th quartile dropped to the median value: $`r as.character(q4.dollars.2015 - q4.projected.2015)`  

Projected revenue decrease calculated by industry: $`r as.character(sum(industry.totals.2015$delta))`   

